### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
spell check on parallel chapter

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo. Considering a domain decomposed
Neumann-Ulam method, we derive an analytic framework based on
algebraic quantities from which to estimate its performance when
applied to neutron transport systems. We then devise a parallel
algorithm for the Neumann-Ulam method based on the multiple-set
overlapping-domain decomposition algorithm and a parallel algorithm
for the MCSA iteration that leverages the parallel Monte Carlo
algorithm and general parallel matrix-vector operations. In identical
fashion to the serial computations presented in
Chapter~\ref{ch:spn_equations}, we use parallel solutions for the
$SP_N$ equations for the fuel assembly criticality problem to verify
the correctness of the parallel algorithm and implementation by
comparing the numerical results against production Krylov
methods. Using the verified implementation of the algorithm, we
perform parallel scaling studies to test its performance on a
leadership class machine and compare this performance to the
production Krylov methods.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions to the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, the solution at each grid point
in the system will be compared to the solution at the same grid point
in the serial reference case with the minimum and maxiumum absolute
differences of these values reported.

@@

From the data we see that for all solvers, the parallel results agree
within the solution tolerance of the reference serial results for all
DOF in the system. In addition, we do not notice any systematic
differences in the results as a function of number of blocks, number
of sets, or amount of overlap. In addition, the parallel GMRES results
have maximum absolute differences of the same magnitude as all MCSA
results and a larger minimum absolute difference than all MCSA results
by three orders of magnitude. Based on these results, MCSA parallized
with an MSOD, and in particular the implementation used here,
generates the correct results with respect to the serial conjugate
gradient reference computation.

@@

To determine the amount of overlap to test for the scaling studies, we
use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
discrete states a history will move on average from its birth site. We
modify Eq~(\ref{eq:step_k_length}) to only account for discrete
movements:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
discrete states a history will move from its birth site. For our
diffusion problem used in these scaling studies, we compute a value of
2.63 states per history for any of the dimensions of the system. This
means that we can expect to end up on average 2.63 states in both the
$i$ and $j$ dimensions away from the starting state. Therefore, we
will use the overlap generation algorithm outlined in
\S~\ref{subsec:msod_generation} such that we parameterically grow the
local domain using 1, 2, 3, and 5 levels to measure how scalability is
affected when this root-mean-squared number of states is
considered. For example, using 3 levels of overlap should keep over
50\% of the histories that would be communicated in the pure domain
decomposition case on-process as this is greater than the
root-mean-squared number of states we would expect a history to
travel.

@@

For both the strong and weak scaling studies, the parallel performance
of the algorithm was either marginally improved or not improved at
all. We would always expect an improvement in performance for using
overlap to prevent the transport of histories from domain to domain
during execution of the Neumann-Ulam algorithm. The reduction in
tranport means a reduction in parallel communication and therefore
should never degrade the performance of the algorithm. It is true that
doing so in fact enhances Monte Carlo performance. However, the
Neumann-Ulam algorithm is not representative of the entire MCSA
algorithm. When overlap is used, an overlapping tally vector is
constructed in parallel such that multiple cores may own multiple
states in the system. When this occurs, the tally data associated with
these shared states must be communicated after the Neumann-Ulam solve
as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
operation then builds the full MCSA correction vector in the
decomposition of the solution vector.

@@

Based on this data, a few levels of overlap within the expected
distance of travel for a history may offer some marginal improvement
in scalability, however, it is not as effective as one might think
given its improvement to reactor physics calculations as observed in
the literature. This is largely due to the density of states in the
system that creates the described performance bottlenecks in the
overlapping tally vector reduction operation. In a Monte Carlo
particle simulation, these overlapping tally reduction operations are
not expected to be peformed at the speed they are required in
MCSA. For example, in a criticality calculation with overlapping
domains, we would expect to perform as many parallel tally vector
reduction in a particle transport simulation as there are particle
stages required for convergence of the eigenvalue. However, these
particle stages will likely be much more costly in terms of wall time
relative to a Neumann-Ulam solve in an MCSA iteration due to the more
extensive particle physics and logic structure that must be
evaluated. Therefore, we can expect this overlapping tally reduction
to be less signficant to scalability in those operations. We do not
have as complicated a logic structure in a history random walk for the
Neumann-Ulam method and therefore overlap can inhibit scalability if
too large.

@@

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative perforamance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. Furthermore, no overlap was required to maintain
iterative performance be letting histories near the boundary take a
few more steps in their random walks. If in the future there are
ill-conditioned problems where using subdomain Monte Carlo reduces the
algorithmic efficiency of MCSA due to the information lost in the
correction vector, overlap would be a viable mechanism to potentially
maintain that iterative performance, keeping in mind the potential
impacts on scalability that were observed in this work.

>>>>

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo. Considering a domain decomposed
Neumann-Ulam method, we derive an analytic framework based on
algebraic quantities from which to estimate aspects of its performance
when applied to neutron transport systems. We then devise a parallel
algorithm for the Neumann-Ulam method based on the multiple-set
overlapping-domain decomposition algorithm and a parallel algorithm
for the MCSA iteration that leverages the parallel Monte Carlo
algorithm and general parallel matrix-vector operations. Using
parallel solutions for the neutron diffusion problem, we verify the
correctness of the parallel algorithm and implementation by comparing
the numerical results against production Krylov methods. With the
verified implementation of the algorithm, we perform parallel scaling
studies to test its performance on a leadership class machine and
compare this performance to the production Krylov methods.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions to the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, the solution at each grid point
in the system will be compared to the solution at the same grid point
in the serial reference case with the minimum and maximum absolute
differences of these values reported.

@@

From the data we see that for all solvers, the parallel results agree
within the solution tolerance of the reference serial results for all
DOF in the system. In addition, we do not notice any systematic
differences in the results as a function of number of blocks, number
of sets, or amount of overlap. In addition, the parallel GMRES results
have maximum absolute differences of the same magnitude as all MCSA
results and a larger minimum absolute difference than all MCSA results
by three orders of magnitude. Based on these results, MCSA parallelized
with an MSOD, and in particular the implementation used here,
generates the correct results with respect to the serial conjugate
gradient reference computation.

@@

To determine the amount of overlap to test for the scaling studies, we
use a variation of Eq~(\ref{eq:step_k_length}) to determine how many
discrete states a history will move on average from its birth site. We
modify Eq~(\ref{eq:step_k_length}) to only account for discrete
movements:
\begin{equation}
  \sqrt{\bar{n^2_k}} = n_s \sqrt{k}\:,
  \label{eq:discrete_distance}
\end{equation}
where $\sqrt{\bar{n^2_k}}$ is now the root-mean-squared number of
discrete states a history will move from its birth site. For our
diffusion problem used in these scaling studies, we compute a value of
2.63 states per history for any of the dimensions of the system. This
means that we can expect to end up on average 2.63 states in both the
$i$ and $j$ dimensions away from the starting state. Therefore, we
will use the overlap generation algorithm outlined in
\S~\ref{subsec:msod_generation} such that we parametrically grow the
local domain using 1, 2, 3, and 5 levels to measure how scalability is
affected when this root-mean-squared number of states is
considered. For example, using 3 levels of overlap should keep over
50\% of the histories that would be communicated in the pure domain
decomposition case on-process as this is greater than the
root-mean-squared number of states we would expect a history to
travel.

@@

For both the strong and weak scaling studies, the parallel performance
of the algorithm was either marginally improved or not improved at
all. We would always expect an improvement in performance for using
overlap to prevent the transport of histories from domain to domain
during execution of the Neumann-Ulam algorithm. The reduction in
transport means a reduction in parallel communication and therefore
should never degrade the performance of the algorithm. It is true that
doing so in fact enhances Monte Carlo performance. However, the
Neumann-Ulam algorithm is not representative of the entire MCSA
algorithm. When overlap is used, an overlapping tally vector is
constructed in parallel such that multiple cores may own multiple
states in the system. When this occurs, the tally data associated with
these shared states must be communicated after the Neumann-Ulam solve
as outlined in \S~\ref{subsec:msod_algorithm}. This reduction
operation then builds the full MCSA correction vector in the
decomposition of the solution vector.

@@

Based on this data, a few levels of overlap within the expected
distance of travel for a history may offer some marginal improvement
in scalability, however, it is not as effective as one might think
given its improvement to reactor physics calculations as observed in
the literature. This is largely due to the density of states in the
system that creates the described performance bottlenecks in the
overlapping tally vector reduction operation. In a Monte Carlo
particle simulation, these overlapping tally reduction operations are
not expected to be performed at the speed they are required in
MCSA. For example, in a criticality calculation with overlapping
domains, we would expect to perform as many parallel tally vector
reduction in a particle transport simulation as there are particle
stages required for convergence of the eigenvalue. However, these
particle stages will likely be much more costly in terms of wall time
relative to a Neumann-Ulam solve in an MCSA iteration due to the more
extensive particle physics and logic structure that must be
evaluated. Therefore, we can expect this overlapping tally reduction
to be less significant to scalability in those operations. We do not
have as complicated a logic structure in a history random walk for the
Neumann-Ulam method and therefore overlap can inhibit scalability if
too large.

@@

From a parallel performance perspective, MCSA may then be best used in
a mode where the Monte Carlo is performed over only the subdomains
with no replication or overlap added to the system. In this mode,
iterative performance is maintained relative to the solutions with
communication even for more ill-conditioned problems with larger
spectral radii. Furthermore, no overlap was required to maintain
iterative performance be letting histories near the boundary take a
few more steps in their random walks. If in the future there are
ill-conditioned problems where using subdomain Monte Carlo reduces the
algorithmic efficiency of MCSA due to the information lost in the
correction vector, overlap would be a viable mechanism to potentially
maintain that iterative performance, keeping in mind the potential
impacts on scalability that were observed in this work.
